QueryMed: An Intuitive SPARQL Query Builder for Biomedical RDF Data

### 2010-02-11 trunk/docs/www2010/paper.tex
merging use case with architecture description

Publicly available data in the biomedical domain has grown significantly over the recent years.   For example, this data include data on drug discovery~\cite{Sharp, Goble}, clinical trial information that are publicly available at ClinicalTrails.gov, disease genes, phenotypes published by the Online Mendelian Inheritance in Man (OMIM), and etc.  With the linked open data movement, the semantic web community has been very proactive in converting these datasets to RDF~\cite{LinkingData}.  In fact, the biomedical domain is among the early successes of adopting semantic web due to the rapidity with which the community has made its data available in RDF triple stores \cite{Yip}.  
%The Semantic Web Health Care and Life Sciences Interest Group (HCLSIG) has been formed with the purpose of exploring the applications of the semantic web to the biomedical domain \cite{HCLSIG}.

Although a plethora of useful biomedical data is currently available in RDF, there is a need for easy-to use systems that do not require the end user to have knowledge of the underlying  structure of the data, and also allow users to run federated queries on multiple SPARQL endpoints.  The state of the art seem to be browsing the data using a linked data browser such as the Tabulator~\cite{Tabulator}.  The linked data approach allows a user to navigate the data in an exploratory manner, but lacks support for efficient filtering and querying of the data in a manner that is amenable for domain experts. Therefore, there is a need to have efficient hybrid interfaces that combine both browsing and querying~\cite{Jentzsch}.

For an example in the biomedical domain, a physician may know her patient's personal information, symptoms, current medications, and genotype. She may wish to determine the patient's treatment plan and identify clinical trials for which the patient is eligible.  Although the physician has a single question--``based on the information I have about this patient, what is the best treatment plan and set of clinical trials available?"--there is no single data source that the physician can use to answer this question. The information that the physician needs must be gathered from numerous data sources such as Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome,  and GO \cite{Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome, GO}. Her question must be broken up into discrete pieces that can be executed individually at one data source at a time.  
Suppose the physician is interested in information on “coronary artery disease” (which she thinks is the cause of her patient's symptoms). The information about the disease can be found in Diseasome, a database that lists diseases, disease genes, and associations between them~\cite{Diseasome}. The physician might also want to know the set of drugs in DailyMed~\cite{DailyMed}, and Drugbank~\cite{Drugbank} that can be used to treat coronary artery disease.  Information on clinical trials that were successful in treating this particular disease of her patient can be found in LinkedCT~\cite{LinkedCT}.

@@

@@

@@

@@

@@

\subsection{A Sample Use Case}

A physician is interested in finding semantic web resources related to coronary artery disease.  She first tries a basic search over all default resources, by entering ``coronary artery disease" in the input search box.   She then sees a list of disease names in the Diseasome database and drugs in DailyMed and Drugbank that relate to coronary artery disease displayed in a table.  She can then filter the results using additional search terms.  For example, she knows that the route of administration of the drug that she is looking for is injection, so she filters the drug query results on the route of administration field using the query term ``injection."  She then prints the table of results.

She is now interested in finding relevant clinical trials for her patient.  However, the clinical trial database LinkedCt is not in the default set of endpoints, so she selects the ``Refine Query" option to choose additional endpoints to search.  She sees a list of default endpoints, and selects the ``Add" option to include an additional endpoint.  After entering the name and URL of the LinkedCT endpoint, she is able to search for clinical trials for which her patient may be eligible.

She is also interested in further refining her search.  She uses the advanced search option to construct a boolean query that takes advantage of the underlying structure of the RDF data in the database.  She searches diseasome for a list of diseases whose class is ``Cardiovascular" or for which the associated gene is ``ABCA1."  Using the QueryMed advanced search interface, the complex SPARQL query corresponding to her question is automatically constructed, and she can view the query results conveniently displayed in a table.

A video illustrating this sample use case can be found at: \\
http://www.youtube.com/watch?v=JCr7vryS9Vg

@@

>>>>

Publicly available data in the biomedical domain has grown significantly over the recent years.   For example, this data include data on drug discovery~\cite{Sharp, Goble}, clinical trial information that are publicly available at ClinicalTrails.gov, disease genes, phenotypes published by the Online Mendelian Inheritance in Man (OMIM), and etc.  With the linked open data movement, the semantic web community has been very proactive in converting these datasets to RDF~\cite{LinkingData}.  In fact, the biomedical domain is among the early successes of adopting semantic web due to the rapidity with which the community has made its data available in RDF triple stores \cite{Yip}.

%The Semantic Web Health Care and Life Sciences Interest Group (HCLSIG) has been formed with the purpose of exploring the applications of the semantic web to the biomedical domain \cite{HCLSIG}.

Although a plethora of useful biomedical data is currently available in RDF, there is a need for easy-to use systems that do not require the end user to have knowledge of the underlying  structure of the data, and that also allow users to run federated queries on multiple SPARQL endpoints.  Many currently available systems are linked data data browsers such as the Tabulator~\cite{Tabulator}, which allow a user to navigate the data in an exploratory manner, but lack support for efficiently filtering and querying the data. Therefore, there is also a  need for efficient hybrid interfaces that allow both browsing and querying~\cite{Jentzsch}.

For an example in the biomedical domain, a physician may know her patient's personal information, symptoms, current medications, and genotype. She may wish to determine the patient's treatment plan and identify clinical trials for which the patient is eligible.  Although the physician has a single question--``based on the information I have about this patient, what is the best treatment plan and set of clinical trials available?"--there is no single data source that the physician can use to answer this question. The information that the physician needs must be gathered from numerous data sources such as Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome,  and GO \cite{Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome, GO}. Her question must be broken up into discrete pieces that can be executed individually at one data source at a time.  
Suppose the physician is interested in information on “coronary artery disease” (which she thinks is the cause of her patient's symptoms). The information about the disease can be found in Diseasome, a database that lists diseases, disease genes, and associations between them~\cite{Diseasome}. The physician might also want to know the set of drugs in DailyMed~\cite{DailyMed}, and Drugbank~\cite{Drugbank} that can be used to treat coronary artery disease.  Information on clinical trial for which her patient may be eligible can be found in LinkedCT~\cite{LinkedCT}.

@@

 For example, a physician interested in finding semantic web resources related to coronary artery disease could use “coronary artery disease”  as a search keyword in the input search box and select the "Query All" option.   The simplicity of the basic query interface allows this end user to rapidly and easily browse semantic web resources of interest.

@@

For example, after searching for "coronary artery disease" in the basic query interface, the physician will see displayed in the Result Browser a list of disease names in the Diseasome database and drugs in DailyMed and Drugbank that relate to coronary artery 
disease. She can then filter the results using additional search terms. For example, she knows that 
the route of administration of the drug that she is looking for is injection, so she filters the drug query results on the route of administration field using the query term “injection.”   The additional filtering capabilities of the Result Browser allows her  to either browse a large number of query results, or refine her search to view the most relevant search terms.

@@

For example, the physician might be interested in finding relevant clinical trials for 
her patient. However, the clinical trial database LinkedCt is 
not in the default set of endpoints, so she selects the “Refine 
Query” option to choose additional endpoints to search. She 
sees a list of default endpoints, and selects the “Add” option 
to include an additional endpoint. After entering the name 
and URL of the LinkedCT endpoint, she is able to search 
for clinical trials for which her patient may be eligible.

@@

For example, the physician might want to further refine her search for cardiovascular diseases. She 
uses the advanced search option to construct a boolean query 
that takes advantage of the underlying structure of the RDF 
data in the database. She searches Diseasome for a list of 
diseases whose class is “Cardiovascular” or for which the as- 
sociated gene is “ABCA1.” Using the QueryMed advanced 
search interface, the complex SPARQL query corresponding 
to her question is automatically constructed (Figure \ref{fig:SPARQL_adv}), and she can 
view the query results conveniently displayed in the Results Browser.

@@

%\subsection{A Sample Use Case}

%A physician is interested in finding semantic web resources related to coronary artery disease.  She first tries a basic search over all default resources, by entering ``coronary artery disease" in the input search box.   She then sees a list of disease names in the Diseasome database and drugs in DailyMed and Drugbank that relate to coronary artery disease displayed in a table.  She can then filter the results using additional search terms.  For example, she knows that the route of administration of the drug that she is looking for is injection, so she filters the drug query results on the route of administration field using the query term ``injection."  She then prints the table of results.

%She is now interested in finding relevant clinical trials for her patient.  However, the clinical trial database LinkedCt is not in the default set of endpoints, so she selects the ``Refine Query" option to choose additional endpoints to search.  She sees a list of default endpoints, and selects the ``Add" option to include an additional endpoint.  After entering the name and URL of the LinkedCT endpoint, she is able to search for clinical trials for which her patient may be eligible.

%She is also interested in further refining her search.  She uses the advanced search option to construct a boolean query that takes advantage of the underlying structure of the RDF data in the database.  She searches diseasome for a list of diseases whose class is ``Cardiovascular" or for which the associated gene is ``ABCA1."  Using the QueryMed advanced search interface, the complex SPARQL query corresponding to her question is automatically constructed, and she can view the query results conveniently displayed in a table.

%A video illustrating this sample use case can be found at: \\
http://www.youtube.com/watch?v=JCr7vryS9Vg

@@

A video illustrating a sample use case can be found at: [USE CASE URL]
